---
title: "Statistical Methods in Epidemiology"
subtitle: "Lab 1"
date: "10/13/2020"
output: github_document
always_allow_html: true
---

## Objectives

The aim of this practical is to display how main measures of disease and effect used in epidemiology are calculated in R.

First, let's load all the libraries that we are going to use. Make sure that you have already installed the respective R packages.

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(gmodels)
library(epiR)
```


## I. Analysis of cohort studies

For this part of the practical we will use data from the Whitehall study. We'll store our data in a *data.frame* object called **wha_data** and create factors for all the categorical variables.

```{r}
wha_data <- read.table("Wha.txt", header = TRUE)
wha_data$group <- factor(wha_data$group, levels = c("1", "2", "3"))
wha_data$work <- factor(wha_data$work, levels = c("high", "low"))
wha_data$smok <- factor(wha_data$smok, levels = c("0", "1", "2", "3"), labels = c("never", "former", "0-19", ">= 20"))
wha_data$all <- factor(wha_data$all, levels = c("1", "0"), labels = c("died", "alive"))
wha_data$IHD <- factor(wha_data$IHD, levels = c("1", "0"), labels = c("IHD", "Other_cause"))

```

### Risk Analysis


Using the `CrossTable()` function from package **gmodels** we can examine the distribution of categorical variables, as well as create contingency tables to examine the relationship between two categorical variables. By default (when we tabulate two variables), `CrossTable()` gives us cell row, column and table proportions, as well as Chi-square contribution for every cell. We can also specify arguments **chisq = TRUE ** and/or **fisher = TRUE** in order for results of a Chi-square test and/or a Fisher's exact test to be included. In this case, we ask from `CrossTable()` just the row and column proportions for every cell and we give more informative labels to the variables that are tabulated. Alternatives to `CrossTable()` are Base's R functions `table()` and `xtabs()`. Note, also, the use of function `with()` that evaluates an expression (here the use of `CrossTable()`) in an environment that is constructed from data that we input (here our data is *data.frame* **wha_data** and `CrossTable()` operates within it).

```{r}
with(wha_data, CrossTable(all))
with(wha_data, CrossTable(IHD, all, prop.t = FALSE, prop.chisq = FALSE, dnn = c("Cause of death", "Status")))
```


In order to assess the risk of death according to working group and calculate the risk ratio of the high grade vs. low grade working groups (as well as some other probably useful measures), we'll use the `epi.2by2()` function from **epiR** package. This function accepts an object of class *table*, which means that we have to create it first in order to proceed (in this case we need a contingency table of status (variable **all**) to working group (variable **work**)) and furthermore we need to specify whether the outcome variable (here variable **all**) lies on rows or columns (argument **outcome = "as.rows"** or **outcome = "as.columns"**). The main argument, however, that needs to be specified is **method**, which defines the study design and consequently determines the measures that will be calculated (Risk Ratio, Rate Ratio or Odds Ratio). As, initially, we will not take into account the fact that we have access to the numbers of years that every person contributed to the study, we will specify the argument **method = TRUE** in order for the risk ratio of the high vs. low grade working groups to be calculated.

```{r}
table1 <- with(wha_data, table(all, work))
epi.2by2(table1, method = "cohort.count", outcome = "as.rows")
```

### Incidence Rate Analysis


In this part we will deploy the years that every person contributed to the study (variable **y**) and we will examine the effect of grade of work on mortality through an incidence rate analysis. For this purpose, we need the total number of deaths and the total person-years in both working groups (high and low). We can extract this information from our initial dataset with the aid of tools from **Tidyverse** package.

```{r message = FALSE, warning = FALSE}
by_work <- wha_data %>%
  group_by(work) %>%
  summarise(Cases = sum(all == "died"), Person_years = sum(y))
by_work
```

To perform the incidence rate analysis, again we are going to use `epi.2by2()` function. However, this time the argument **method = "cohort.time"** has to be specified. Also, in this case the function accepts a *table* with the number of cases in the first column, the number of person-years in the second and the rows grouped by the exposure (here variable **work**), and that's why we created the *data.frame* **by_work**. Our last act before the use of `epi.2by2()` is to extract the last two columns from **by_work** (because we just want the number of cases and the person-years) and create a *table* from them.

```{r message = FALSE, warning = FALSE}
table2 <- as.table(as.matrix(by_work[,2:3]))
epi.2by2(table2, method = "cohort.time")
```


Afterwards, we'd like to examine the effect of smoking status (variable **smok**) on the risk of death. Smoking status is a categorical variable with 4 levels. For exposures with more than 2 levels `epi.2by2` does not work, but we can manual calculate the point estimates for incidence rates for all the levels. In order to have a picture of how rates are modified among levels of smoking status and if they have a statistically significant difference we will create a plot of these point estimates along with their 95% confidence limits. To calculate the confidence limits, we will use the normal approximation to the distribution of log of rates.

```{r message = FALSE, warning = FALSE}
by_smok <- wha_data %>%
  group_by(smok) %>%
  summarise(Cases = sum(all == "died"), Person_years = sum(y)) %>% 
  mutate(Inc_Rate = 1000 * Cases/Person_years, 
         lower_l = Inc_Rate/exp(1.96 * sqrt(1/Cases)), 
         upper_l = Inc_Rate*exp(1.96 * sqrt(1/Cases)))

by_smok

ggplot(by_smok, aes(smok, Inc_Rate)) + 
  geom_point(size = 3) + 
  geom_errorbar(aes(ymin = lower_l, ymax = upper_l), color = "red", width = 0.5) +
  ylab("Incidence Rates by 1000 person-years") +
  xlab("Smoking")
```


Another approach to the examination of the effect of smoking to the risk of death is to split our sample to smokers and non-smokers based on the information of variable **smok** and perform again an incidence rate analysis. 

```{r message = FALSE, warning = FALSE}
wha_data <- wha_data %>% 
  mutate(smok2 = factor(ifelse(smok %in% c("never", "former"), 0, 1), levels = c("1", "0"), labels = c("smokers", "non_smokers")))

by_smok2 <- wha_data %>%
  group_by(smok2) %>% 
  summarize(Cases = sum(all == "died"), Person_years = sum(y))

table3 <- as.table(as.matrix(by_smok2[,2:3]))

epi.2by2(table3, method = "cohort.time") 


```


## II. Analysis of case-control studies

For this part of the practical we will use data from the case-control study of HIV infection in Mwanza, Tanzania. We'll store our data in an object of class *data.frame* called **mw_data** and create factors for our categorical variables. Note that variable **skin** has one missing value (9) and it will be converted to *NA* by not including it to the levels of the factor.

```{r}
mw_data <- read.table("mwanza.txt", header = TRUE)
mw_data$case <- factor(mw_data$case, levels = c("1", "0"), labels = c("Cases", "Controls"))
mw_data$ed <- factor(mw_data$ed, levels = c("0", "1", "2", "3"), labels = c("none", "elementary", "medium", "high"))
mw_data$skin <- factor(mw_data$skin, levels = c("1", "0"), labels = c("yes", "no"))
mw_data$work <- factor(mw_data$work, levels = c("1", "2"))
```


First, we'll examine the difference in education levels in cases and controls.

```{r}
with(mw_data, CrossTable(case, ed, expected = FALSE, prop.r = TRUE, prop.c = TRUE, prop.t = FALSE, prop.chisq = FALSE, dnn = c("HIV", "Education")))
```

Then, we'll calculate the odds of HIV according to education level. As education has more than 2 levels, we'll do it manually. We'll also calculate 95% confidence limits for odds based on normal approximation of log of odds.

```{r message = FALSE, warning = FALSE}
by_educ <- mw_data %>% 
  group_by(ed) %>% 
  summarise(Cases = sum(case == "Cases"), Controls = sum(case == "Controls")) %>% 
  mutate(Odds = Cases/Controls, 
         lower_l = Odds/exp(1.96*sqrt((1/Cases)+(1/Controls))), 
         upper_l = Odds*exp(1.96*sqrt((1/Cases) + (1/Controls))))

by_educ
```

Next, we'd like to examine whether the odds of HIV vary according to whether the woman has skin incisions or tattoos (variable **skin**). Skin has two levels, so, once more, `epi.2by2()` will be our main tool. Since our data are from a case-control study, we need to specify the argument **method = "case.control"**. We remind that this function accepts an object of class *table*.

```{r}
table4 <- with(mw_data, table(case, skin))

epi.2by2(table4, method = "case.control", outcome = "as.rows")
```

Finally, using the variable of education levels (**ed**), we will create the variable **ed2** that shows whether a woman has no formal education or some education and perform an odds analysis.

```{r}
mw_data <- mw_data %>% 
  mutate(ed2 = factor(ifelse(ed == "none", 0, 1), levels = c("1", "0"), labels = c("some_education", "no_education")))

table5 <- with(mw_data, table(case, ed2))

epi.2by2(table5, method = "case.control", outcome = "as.rows" )

```

