---
title: 'Statistical Methods in Epidemiology'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    pandoc_args: --webtex=https://latex.codecogs.com/png.latex?%5Cdpi{150}
always_allow_html: true
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999, digits = 3)
```


<p style="text-align:center";>
  <b> 
    <font size="6">Lab 3</font>
  </b>
</p>

## Table of Contents
1. [Analyzing Rates](#rates)
2. [Exposures with More Than Two Levels](#exposure)
3. [Controlling for a Confounding Variable](#confounding)
4. [Likelihood Ratio and Wald Tests](#wald)
5. [Metric Exposures and Linear Effects](#explin)
6. [Testing for Linearity](#linearity)
7. [Interaction](#interaction)


In the practical that follows, *diet* data set, which contains data from a pilot study of 337 men who kept a record of their fully weighted diet over two weeks, will be analyzed. The particular data set lies in **biostat3** package, so we will load it along with the rest of packages that are going to be used.

```{r, message = FALSE, warning = FALSE}
library(biostat3)
library(tidyverse)
library(kableExtra)
library(lmtest)
```

A brief investigation of our data reveals that variable **energy** is probably measured in calories, so we will convert it into kcals in order to obtain the same results with the respective **stata** practical. Also, we will create the categorical variables **eng3** and **htgrp** that were also created in the previous lab and will be used later on.

```{r}
data(diet)

diet <- diet %>%
  mutate(energy = energy / 1000, 
         eng3 = cut(energy, breaks = c(1.5, 2.5, 3, 4.5), labels = c("0", "1", "2")), 
         htgrp = cut(height, breaks = c(150, 170, 175, 180, 195), labels = c("0", "1", "2", "3")))
```

Preview of the variables created below:

```{r, message = FALSE, warning = FALSE, echo = FALSE}
diet %>%
  dplyr::select(energy, eng3, htgrp) %>%
  head() %>% 
  kbl() 
```

In what follows, we will mainly work with Poisson regression models. As, on most occasions, we prefer the <u>incidence rate ratios</u> to effects on a log scale, we will create a function that accepts a Poisson model as argument and returns the incidence rate ratio for each of the independent variables along with their standard error (calculated using delta method), p-value and 95% confidence intervals (see also **mfx** package for something similar). 

<p style="color:red";><b>Don't be discouraged if you don't understand. The following requires enough practice with R and programming concepts in general. This function is for the authors' ease to continue with the Lab. The main idea for the creation of a function is to not repeat the same code chunks across the script.</b></p>

```{r}
r_table <- function(model) {
  if (is(model, "glm") == TRUE) {
    if (family(model)$family == "poisson") {
      modification <- 
        model %>% 
        summary() %>% 
        coefficients() %>% 
        as.data.frame() %>% 
        slice(-1) %>% 
        rownames_to_column(var = "Predictors") %>% 
        mutate(IRR = exp(Estimate), 
               Std_Err = IRR * `Std. Error`, 
               pval = sprintf("%1.3f", `Pr(>|z|)`),
               pvalue = ifelse(pval < 0.001, '<0.001', pval)) %>%
        select(Predictors, 
               IRR, Std_Err, 
               "z_value" = `z value`, 
               "P-Value" = pvalue)
      
      if (nrow(modification) > 1) {
        final_table <- bind_cols(modification, 
                                 LL = exp(confint.default(model)[-1, ])[, 1], 
                                 UL = exp(confint.default(model)[-1, ])[, 2])
        } else {
          final_table <- bind_cols(modification, 
                                   LL = exp(confint.default(model)[-1, ])[1], 
                                   UL = exp(confint.default(model)[-1, ])[2])
        }
      } else if (family(model)$family == "binomial") {
              modification <- 
        model %>% 
        summary() %>% 
        coefficients() %>% 
        as.data.frame() %>% 
        slice(-1) %>% 
        rownames_to_column(var = "Predictors") %>% 
        mutate(OR = exp(Estimate), 
               Std_Err = OR * `Std. Error`, 
               pval = sprintf("%1.3f", `Pr(>|z|)`),
               pvalue = ifelse(pval < 0.001, '<0.001', pval)) %>%
        select(Predictors, 
               OR, Std_Err, 
               "z_value" = `z value`, 
               "P-Value" = pvalue)
      
      if (nrow(modification) > 1) {
        final_table <- bind_cols(modification, 
                                 LL = exp(confint.default(model)[-1, ])[, 1], 
                                 UL = exp(confint.default(model)[-1, ])[, 2])
        } else {
          final_table <- bind_cols(modification, 
                                   LL = exp(confint.default(model)[-1, ])[1], 
                                   UL = exp(confint.default(model)[-1, ])[2]) 
        } 
      } else {
        stop("Family of the model is not Poisson or Binomial")
        }
    } else {
      stop("Object is not of class glm")
      } 
    return(final_table) 
}
```

## I. Analyzing Rates <a name="rates"></a>

We will investigate the effect of energy intake on the rate of cases with CHD fitting a poisson regression model with `glm()` function, specifying the argument **family = "poisson"** (by default, when **family = "poisson"** is specified, link function is the natural logarithm). The version of energy intake that we will originally use is the categorical one (variable **hieng**). Note, that person-years at risk for each subject are already in our data set (variable **y**) and we will introduce them as an offset into our model (argument **offset**). Estimated coefficients along with further useful information about the model can be provided with Base R function `summary()` or `summ()` from **jtools** package. We will also use the function that we created above in order to obtain the incidence rate ratio of CHD comparing men with high energy intake to those with low.

```{r}
fit1 <- glm(chd ~ hieng, data = diet, family = "poisson", offset = log(y))

r_table(fit1) %>%
  kbl(digits = 2) %>%
  kable_styling()
```


## II. Exposures with More Than Two Levels <a name="exposure"></a>

As before, we 'll investigate the effect of energy intake on the rate of cases with CHD. This time, however, we 'll use the second categorical version of energy, **eng3**.

```{r}
fit2 <- glm(chd ~ eng3, data = diet, family = "poisson", offset = log(y))

r_table(fit2) %>%
  kbl(digits = 2) %>%
  kable_styling()

```


## III. Controlling for a Confounding Variable <a name="confounding"></a>

Now, we'll find the effect of high energy (variable **hieng**) controlled for **job**, as well as the effect of **eng3** controlled for: 

1. **job** 
2. **job** and **htgrp**.

```{r}
fit3 <- glm(chd ~ hieng + job, data = diet, family = "poisson", offset = log(y))

r_table(fit3) %>%
  kbl(digits = 2) %>%
  kable_styling()


fit4 <- glm(chd ~ eng3 + job, data = diet, family = "poisson", offset = log(y))

r_table(fit4) %>%
  kbl(digits = 2) %>%
  kable_styling()


fit5 <- glm(chd ~ eng3 + job + htgrp, data = diet, family = "poisson", offset = log(y))

r_table(fit5) %>%
  kbl(digits = 2) %>%
  kable_styling()
```


## IV. Likelihood Ratio and Wald Tests <a name="wald"></a>

We 'll perform a likelihood ratio test to test the hypothesis that coefficients of **eng3** from model *fit2* are zero. We 'll present two ways: 

1. the manual using `logLik()` function, from Base R, that calculates the log-likelihood of a model
2. the automatic using `lrtest()` function from **lmtest** package.
           
In both cases, initially, we need to create the reduced model (i.e. the model without the variables whose coefficients are tested to be equal to zero). We 'll accomplish it using `update()` function. 

```{r}
fit6 <- update(fit2, .~.-eng3)

pchisq(q = -2 * (as.numeric(logLik(fit6)) - as.numeric(logLik(fit2))), df = 2, lower.tail = FALSE)

lrtest(fit2, fit6) %>%
  kbl(digits = 3) %>%
  kable_styling()
```

An approximation of likelihood ratio test, the Wald test, can be implemented with `waldtest()` function, again from **lmtest** package, specifying the argument **test = "Chisq"**.

```{r}
waldtest(fit2, fit6, test = "Chisq") %>%
  kbl(digits = 4) %>%
  kable_styling()
```


5. Same as before, but this time we create a model with **hieng**.

```{r}
hiengfit <- glm(chd ~ hieng + job, data = diet, family = "poisson", offset = log(y))

nohiengfit <- glm(chd ~  job, data = diet, family = "poisson", offset = log(y))

lrtest(hiengfit, nohiengfit) %>%
  kbl(digits = 3) %>%
  kable_styling()
```

6. Testing the addition of **eng3** in the model that contains already the **job**.

```{r}
eng3fit <- glm(chd ~ eng3 + job, data = diet, family = "poisson", offset = log(y))

noeng3fit <- glm(chd ~ job, data = diet, family = "poisson", offset = log(y))

lrtest(eng3fit, noeng3fit) %>%
  kbl(digits = 3) %>%
  kable_styling()
```

## V. Metric Exposures and Linear Effects <a name="explin"></a>

```{r}
fit8 <- glm(chd ~ height, data = diet, family = "poisson", offset = log(y))

fit8 %>%
  r_table() %>%
  kbl(digits = 3) %>%
  kable_styling()
```

For the model with **fat** as predictor, we will use the *diet.dta* dataset that we used in class. The `*biostat3::diet*` does not include fat as a variable in the dataset. 

```{r}
diet2 <- haven::read_dta('diet.dta')

# The diet2 dataset does not contain the person-years, so we have to create it.
diet2 <- diet2 %>%
  mutate(y = lubridate::time_length(difftime(dox, doe), 'years'))

fatfit <- glm(chd ~ fat, data = diet2, family = "poisson", offset = log(y))

fatfit %>%
  r_table() %>%
  kbl(digits = 4) %>%
  kable_styling()
```

For every 1 g/day increase in fat we expect on average 2% decrease in the CHD mortality rate. But what if we want the 10 g/day increase? All we have to do is isolate the IRR and raise it to the tenth power.

```{r}
fatfit %>%
  r_table() %>%
  .[['IRR']] %>%
  . ^ 10
```

So, for every 10 g/day increase in fat we expect on average 18.4% decrease in CHD mortality rate.

## VI. Testing for Linearity <a name="linearity"></a>

First of, we are going to create the **htsq** variable, that is defined as $htsq = height^2$. Next, we will proceed with the model that contains **htsq** and **height** as predictors.

```{r}
diet2 <- diet2 %>%
  mutate(htsq = height ^ 2)
```

```{r}
htsqfit <- glm(chd ~ height + htsq, data = diet2, family = "poisson", offset = log(y))

# Extract the coefficients of the model above
res <- htsqfit %>% 
  summary() %>% 
  .$coef

# Combine it with the 95% CIs
cbind(res, confint.default(htsqfit)) %>% 
  kbl(digits = 3) %>%
  kable_styling()
```

Notice that from the Wald test we get $z ^ 2 = 0.281, p = 0.596 > 0.05$ for the **htsq** term.

Another way of doing the same as above:

Combining *fit8* and *htsqfit*, we want to evaluate the addition of the squared term with the Likelihood ratio test. 

```{r}
lrtest(fit8, htsqfit) %>%
  kbl(digits = 3) %>%
  kable_styling()
```

From the table above, we observe that $p = 0.583 > 0.05$, hence the squared term for the height is not needed.

## VII. Interaction <a name="interaction"></a>

We are going to investigate the interaction between **hieng** and **job** and its statistical significance.

```{r}
inter1 <- glm(chd ~ job * hieng, data = diet, family = "poisson", offset = log(y))

nointer1 <- glm(chd ~ job + hieng, data = diet, family = "poisson", offset = log(y))

lrtest(inter1, nointer1) %>%
  kbl(digits = 3) %>%
  kable_styling()
```

We conclude that, the addition of the interaction terms is not statistically significant in the 5% confidence level since $X^2 = 0.33 (df = 2), p = 0.847$.

Hence, the effect of high energy on the rates of CHD mortality are not different within the different levels of job.

Next, we will investigate the interaction between **hieng** and **htgrp**.

```{r}
inter2 <- glm(chd ~ hieng * htgrp, data = diet, family = "poisson", offset = log(y)) 

nointer2 <-  glm(chd ~ hieng + htgrp, data = diet, family = "poisson", offset = log(y))

lrtest(inter2, nointer2) %>%
  kbl(digits = 3) %>%
  kable_styling()
```

From the likelihood ratio test, we infer that $X^2 = 0.947(df = 3), p = 0.814$

The above results, show that the effect of high energy on the rates of CHD mortality are not different within the different levels of height groups.

Finally, we are going to test the interaction between **hieng** and **height**.

```{r}
inter3 <- glm(chd ~ hieng * height, data = diet, family = "poisson", offset = log(y)) 

nointer3 <- glm(chd ~ hieng + height, data = diet, family = "poisson", offset = log(y)) 

lrtest(inter3, nointer3) %>%
  kbl(digits = 3) %>%
  kable_styling()
```

Same as before, we do not have enough enough evidence to conclude that interaction between the variables mentioned above is statistically significant in the 5% confidence level, since $X^2 = 0.388(df = 1), p = 0.533$.